---
Title: Course Project
Author: Basemah Alshemali
Date: 12/12/2020
Coursera course: Practical Machine Learning
Goal of this project: to use data from accelerometers on the belt, forearm, arm, and
  dumbel to determine how well a participant was performing barbell lifts. To provide
  input information, six participants were asked to perform barbell lifts correctly
  and incorrectly in 5 different ways.
More information is available from the website here: http://groupware.les.inf.puc-rio.br/har
  (see the section on the Weight Lifting Exercise Dataset).
output: pdf_document
---


# Plan

At first, we learn about the data and find out which data can be useful for prediction and which data cleaning must be done to make the data usable.

The plan is as follows:
1- Explore the data and apply necessary data cleaning.
2- Train three algorithms and evaluate their performance on the validation data set.
3- Select the best performing one based on the accuracy.
4- Run the prediction on the testing data set.



# Required libraries
To run this project, the following libraries are required.

```{r}
library(readr)
library(dplyr)
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(corrplot)
library(rattle)
```

# Loading data
We load the training and testing data to "training" and "testing" streams.

```{r, warning=FALSE}
training <- read_csv("pml-training.csv")
testing <- read_csv("pml-testing.csv")
```

# Exploring data

# Exploring the dimensions of data

```{r}
dim(training)
dim(testing)
```

We can see that both training and testing files have 160 columns. The training set has 19622 rows, while the testing set has 20 rows.


# Returning the first part of the data

```{r}
head(training)
```
The training set contains a number of columns with a large number of missing (NA) values. 
We will remove columns with over 90% NA’s from both the training and testing sets as they don't add any value for model prediction. 
In addition, the first seven columns will be removed since they are either ID variables or date/time stamp related.


# Cleaning data

```{r}
training <- training[,-c(1:7)]
testing <- testing[,-c(1:7)]

varsNA <- sapply(training, function(x) mean(is.na(x))) > 0.9
training <- training[,varsNA==FALSE]
testing  <- testing[,varsNA==FALSE]
```


```{r}
dim(training)
dim(testing)
```
As we can see, we have left with 53 columns after the cleaning.


# Spliting the training set into a sub-training (70%) and testing(30%) set

```{r}
set.seed(1)
inTrain <- createDataPartition(y=training$classe, p=0.7, list=FALSE)
training_sub <- training[inTrain,]
testing_sub <- training[-inTrain,]
```

# Building models

We will use 3 methods and choose the one having the best accuracy to predict the outcome variable in the testing set. A confusion matrix plotted at the end of each model will help visualize the analysis better. The following methods will be evaluated:
1. Ssupport vector machine (SVM)
2. Gradient Boosted Machines (GBM)
3. Random Forests (RF)


# Support vector machine (SVM)

# Training model

```{r}
set.seed(1)
modFit1<-train(classe~.,method="svmRadial",data=training_sub)
```

# Estimating the performance of the SVM on the validation data set.

```{r}
predictSVM <- predict(modFit1, testing_sub)
```

# Confusion matrix

```{r}
confusionMatrix(table(testing_sub$classe, predictSVM))
```


## Visualization

```{r}
plot(predictSVM, main="SVM Prediction", ylab="Samples", xlab="Classes")
```


# Gradient Boosted Machines (GBM)

# Training model
```{r}
modFit2<-train(classe~.,method="gbm",data=training_sub)
```

# Estimating the performance of the GBM on the validation data set.

```{r}
predictGBM <- predict(modFit2, testing_sub)
```

# Confusion matrix

```{r}
confusionMatrix( table(testing_sub$classe, predictGBM))
```

# Visualization

```{r}
plot(predictGBM, main="GBM Prediction", ylab="Samples", xlab="Classes")
```


# Random Forests (RF)

# Training model

```{r}
modFit3<-train(classe~.,method="rf",data=training_sub)
```

# Estimating the performance of the RF on the validation data set.

```{r}
predictRF <- predict(modFit3, testing_sub)
```

# Confusion matrix

```{r}
confusionMatrix( table(testing_sub$classe, predictRF ) )
```

## Visualization

```{r}
plot(predictRF, main="RF Prediction", ylab="Samples", xlab="Classes")
```


# Predicting for testing set

The Random Forests has the best accuracy of 99.13%, therefore, this model will be used on the testing data.

```{r}
pred <- predict(modFit3, testing)
```

# Prediction of the 20 cases

```{r}
pred
```



